Executive Constraints and Civil Conflict Onset

### 2020-08-30 paper.Rmd
Manuscript edits

---
title: Executive Constraints and Civil Conflict Onset
author:
    name: J.L.A. Krusell
    affiliation: Department of Peace and Conflict Research
    supervisor: Kristine Eck
    degree: "Master's Thesis"
date: "Spring 2020"
titlelogo: |
    ```{=latex}
    assets/UU_logo_CMYK-eps-converted-to.pdf
    ```
wordcount: 15000
suppress-bibliography: false
bibliography: library.bib
csl: assets/chicago-author-date.csl
link-citations: true
documentclass: scrreprt
classoption: paper=letter
geometry:
    - top=30mm
    - heightrounded
fontsize: 12pt
linestretch: 1.5
header-includes:
    - \usepackage{amsmath}
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
    - \BeforeBeginEnvironment{align*}{\begin{singlespace}\vspace*{-\baselineskip}}
    - \AfterEndEnvironment{align*}{\end{singlespace}\noindent\ignorespaces}
    - \counterwithout{table}{chapter}
    - \counterwithout{figure}{chapter}
    - \usepackage[font={small}]{caption}
output:
    pdf_document:
        fig_caption: true
        keep_tex: true
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--template", "assets/thesis_template.latex",
                      "--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate",
                      "--lua-filter", "utils/text_refs/text_refs.lua"]
    html_document:
        smart: true
        toc: true
abstract: |
    **Abstract**: Do institutional constraints on a regime's executive decrease the
    likelihood of civil conflict onset? An unconstrained executive is free
    to loot state resources, undermining the state's capacity to
    effectively deal with nascent rebellions, and may be more likely to
    engage in violent repression, especially in the form of indiscriminate
    violence. This can encourage political grievances, lead to a loss of
    legitimacy for the regime, and provide the opportunity for rebel
    groups to attract new members by offering protection against
    government violence. Furthermore, the lack of guaranteed checks to
    executive power may incentivise actors to avoid bargained solutions if
    they fear future extralegal retributions. Taken together, where there
    is a lack of effective constraints against the executive there may be
    an increased risk for the onset of civil conflict. To test this
    proposition I employ a bayesian latent variable model which estimates
    executive constraint as a latent factor derived from several manifest
    variables and relates it to a binary measure of civil conflict onset
    through a logistic regression.  The primary finding is that there is a
    negative relationship such that the predicted probability for civil
    conflict is lower where there exists higher levels of executive
    constraints. However, this is conditional on the level of GDP per
    capita; for low-income country-years the relationship between
    executive constraints and civil conflict onset is indeterminate
    possibly because it is easier to recruit and mobilize fighters
    regardless of the overall level of executive constraints. The issue of
    poor model fit means that the presented results should be considered
    tentative at best.
---

@@

options(knitr.kable.NA = ".")
opts_chunk$set(echo = F, message = F, warning = F,
               fig.width = 5, fig.height = 3, cache = T)
```

@@

## Political Institutions and Civil Conflict

@@

[^model]: The model code is written in Stan [@Carpenter2017] and was
    run using the `CmdStan` interface version 2.23.0 [@Stan2020] with
    four parallel chains at 2000 iterations each with a warmup
    of 1000. Convergence was assessed according to the potential scale
    reduction factor $\widehat{R}$ and effective sample size (ESS)
    [@Vehtari2019] in addition to standard Stan diagnostics. All
    pre/post-processing was done using R 3.6.2 [@Core2019]. The Stan
    model code is available in [Appendix A](#appendix-a) and the full
    project in the form of a reproducible pipeline, including the
    Rmarkdown source for this manuscript, is available at
    [https://github.com/jsks/ma-thesis](https://github.com/jsks/ma-thesis).
    Finally, a simulation analysis assessing the model and it's
    ability to recover simulated true values is available in [Appendix
    B](#appendix-b).

@@

q <- summary_stats(df)

@@

```{r results = "hide", fig.width = 8, fig.cap = "\\label{fig:map} (ref:map)"}
map_data <- st_read("data/raw/cshapes_0.6/cshapes.shp") %>%
    filter(GWEYEAR == max(GWEYEAR)) %>%
    select(country_name = CNTRY_NAME, gwid = GWCODE)

@@

tunisia_idx <- filter(final.df, country_name == "Tunisia", year %in% c(2000, 2010)) %$%
    reduced_idx

tunisia_p <- signif(mean(theta[, tunisia_idx[1]] < theta[, tunisia_idx[2]]) * 100, 2)
```

Finally, to expand upon the country selection from figure
\ref{fig:theta} and provide a temporal comparison, figure
\ref{fig:tie} plots the $\theta$ estimates for 2010 versus 2000 by
country. A full list of all countries is available in [Appendix
C](#appendix-c). Note, the statistical significance of differences
between years for a single country cannot strictly be judged by
overlapping credible intervals [@Schenker2001]. For example, Chile,
despite having overlapping credible intervals, still has a posterior
probability of approximately $`r chile_p`$ percent that the estimate
for 2000 is lower than 2010. In the case of Tunisia with even more
visual overlap, there is a $`r tunisia_p`$ percent posterior
probability that the estimate for 2010 is higher than 2000.

@@

beta <- extract(reg, "beta")
colnames(beta) <- c("exec_constraints", colnames(X),
                    sprintf("exec * %s", colnames(X)[interaction_idx]))

@@

```{r results = "hide"}
argentina <- filter(merged.df, country_name == "Argentina", year == 1989) %$%
    prettyNum(rgdpepc, big.mark = ",")

slovakia <- filter(merged.df, country_name == "Slovakia", year == 2003) %$%
    prettyNum(rgdpepc, big.mark = ",")

cmb <- filter(merged.df, country_name == "Cambodia", year == 1987) %>%
    select(pop, area_sqkm) %>%
    mutate(area_sqkm = signif(area_sqkm, 3),
           pop = signif(pop, 3))

india <- filter(merged.df, country_name == "India", year == 1976) %>%
    select(pop, area_sqkm, meanelev) %>%
    mutate(pop = signif(pop, 3),
           area_sqkm = signif(area_sqkm, 3),
           meanelev = signif(area_sqkm, 3))

georgia <- filter(merged.df, country_name == "Georgia", year == 2012) %$%
    signif(meanelev, 3)

russia <- filter(merged.df, country_name == "Russia", year == 2012) %$%
    signif(meanelev, 3)

thai <- filter(merged.df, country_name == "Thailand", year == 2002) %$%
    { discrimpop * 100 }

bhutan <- filter(merged.df, country_name == "Bhutan", year == 1989) %$%
    { discrimpop * 100 }

@@

alpha_s <- quantile(alpha, probs = c(0.1, 0.9)) %>% signif(3)
alpha_interval <- sprintf("[%g,%g]", alpha_s[1], alpha_s[2])
```

@@

```{r results = "hide"}
gdp_rwa <- filter(merged.df, country_name == "Rwanda", year == 2007) %$%
    rgdpepc %>%
    prettyNum(big.mark = ",")

gdp_phl <- filter(merged.df, country_name == "Philippines", year == 2008) %$%
    rgdpepc %>%
    prettyNum(big.mark = ",")

gdp_po <- filter(merged.df, country_name == "Portugal", year == 2009) %$%
    rgdpepc %>%
    prettyNum(big.mark = ",")
```

@@

## A - Stan Code {#appendix-a}

@@

## B - Prior Simulation {#appendix-b}

In order

```{r eval = F}
prior_plot <- function(range, fn, title = NULL, ...) {
    data.frame(x = range) %>%
        ggplot(aes(x)) +
        stat_function(fun = fn, ...) +
        ggtitle(label = title) +
        theme(axis.title.y = element_blank(),
              axis.text.y = element_blank(),
              axis.ticks.y = element_blank())
}

p1 <- prior_plot(c(-20, 20), dnorm, args = list(mean = 0, sd = 5),
                 title = "Normal(0, 5)")
p2 <- prior_plot(c(-10, 10), dnorm, args = list(mean = 0, sd = 2.5),
                 title = "Normal(0, 2.5)")
p3 <- prior_plot(c(0, 5), dlnorm, args = list(meanlog = 0, sdlog = 0.5),
                 title = "Log-Normal(0, 0.5)")
p4 <- prior_plot(c(0, 3), dweibull, args = list(shape = 5, scale = 1),
                 title = "Weibull(5, 1)")
p5 <- prior_plot(c(0, 10), dhcauchy, args = list(sigma = 1),
                 title = "HalfCauchy(0, 1)")
p6 <- prior_plot(c(0, 1), dbeta, args = list(shape1 = 5, shape2 = 5),
                 title = "Beta(5, 5)")

grid.arrange(p1, p2, p3, p4, p5, p6, ncol = 2, nrow = 3)

```{r eval = F, warning = F}
true_values <- readRDS("posteriors/sim/simulated_parameters.rds")

fa <- read_post("posteriors/sim/fa_posteriors.csv")
reg <- read_post("posteriors/sim/reg_posteriors.csv")

fit <- bind_cols(fa, reg)

post <- post_summarise(fit, c("lambda", "psi", "gamma", "delta",
                              "kappa", "rho", "eta", "alpha", "beta",
                              "sigma")) %>%
    rename(codelow = `2.5%`, median = `50%`, codehigh = `97.5%`) %>%
    mutate(type = "Posterior Estimate")

pars.df <- filter(true_values, !grepl("theta", parameter)) %>%
    bind_rows(post) %>%
    mutate(type = as.factor(type))

ggplot(pars.df, aes(parameter, point, color = type)) +
    geom_point() +
    geom_errorbar(aes(parameter, ymin = codelow, ymax = codehigh),
                  width = 0) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```{r eval = F, warning = F}
latent.df <- post_summarise(fit, pars = "theta", probs = c(0.025, 0.975)) %>%
    rename(codelow = `2.5%`, codehigh = `97.5`)

x <- mean(theta <= latent.df$codehigh & theta >= latent.df$codelow)

>>>>

---
title: Executive Constraints and Civil Conflict Onset
author:
    name: J.L.A. Krusell
    affiliation: Department of Peace and Conflict Research
    supervisor: Kristine Eck
    degree: "Master's Thesis"
date: "Spring 2020"
titlelogo: |
    ```{=latex}
    assets/UU_logo_CMYK-eps-converted-to.pdf
    ```
wordcount: 15000
suppress-bibliography: false
bibliography: library.bib
csl: assets/chicago-author-date.csl
link-citations: true
documentclass: scrreprt
classoption: paper=letter
geometry:
    - top=30mm
    - heightrounded
fontsize: 12pt
linestretch: 1.5
header-includes:
    - \usepackage{amsmath}
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
    - \BeforeBeginEnvironment{align*}{\begin{singlespace}\vspace*{-\baselineskip}}
    - \AfterEndEnvironment{align*}{\end{singlespace}\noindent\ignorespaces}
    - \counterwithout{table}{chapter}
    - \counterwithout{figure}{chapter}
    - \usepackage[font={small}]{caption}
output:
    pdf_document:
        fig_caption: true
        keep_tex: true
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--template", "assets/thesis_template.latex",
                      "--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate",
                      "--lua-filter", "utils/text_refs/text_refs.lua"]
    html_document:
        smart: true
        toc: true
abstract: |
    **Abstract**: Do institutional constraints on a regime's executive decrease the
    likelihood of civil conflict onset? An unconstrained executive is free
    to loot state resources, undermining the state's capacity to
    effectively deal with nascent rebellions, and may be more likely to
    engage in violent repression, especially in the form of indiscriminate
    violence. This can encourage political grievances, lead to a loss of
    legitimacy for the regime, and provide the opportunity for would be rebel
    groups to attract new members by offering protection against
    government violence. Furthermore, the lack of guaranteed checks to
    executive power may incentivise actors to avoid bargained solutions if
    they fear future extralegal retributions. Taken together, where there
    is a lack of effective constraints against the executive there may be
    an increased risk for the onset of civil conflict. To test this
    proposition I employ a bayesian latent variable model which estimates
    executive constraint as a latent factor derived from several manifest
    variables and relates it to a binary measure of civil conflict onset
    through a logistic regression.  The primary finding is that there is a
    negative relationship such that the predicted probability for civil
    conflict is lower where there exists higher levels of executive
    constraints. However, this is conditional on the level of GDP per
    capita; for low-income country-years the relationship between
    executive constraints and civil conflict onset is indeterminate
    possibly because it is easier to recruit and mobilize fighters
    regardless of the overall level of executive constraints. The issue of
    poor model fit means that the presented results should be considered
    tentative at best.
---

@@

options(knitr.kable.NA = ".")
opts_chunk$set(echo = F, message = F, warning = F,
               fig.width = 5, fig.height = 3, cache = F)

variable_names <- c("rgdpepc" = "GDPpc (ln)",
                    "rgdpepc_gro" = "GDPpc Growth",
                    "cinc" = "CINC",
                    "discrimpop" = "Discriminated Pop.",
                    "pop_density" = "Pop. Density (ln)",
                    "meanelev" = "Mean Elevation (ln)",
                    "rlvt_groups_count" = "Relevant Groups",
                    "neighbour_conflict" = "Neighbouring Conflict",
                    "major_ongoing" = "High Int. Conflict",
                    "independence" = "Independence",
                    "exec_constraints" = "Exec. Constraints",
                    "exec * rgdpepc" = "Exec. Constraints * GDPpc")

# Don't include this in thesis.utils since the variable names are hardcoded.
translate <- function(x) {
    variable_names[match(x, names(variable_names))] %>% unname
}
```

@@

## Political Institutions

@@

[^model]: The model code is written in Stan [@Carpenter2017] and was
    run using the `CmdStan` interface version 2.23.0 [@Stan2020] with
    four parallel chains at 2000 iterations each with a warmup
    of 1000. Convergence was assessed according to the potential scale
    reduction factor $\widehat{R}$ and effective sample size (ESS)
    [@Vehtari2019] in addition to standard Stan diagnostics. All
    pre/post-processing was done using R 3.6.2 [@Core2019]. The Stan
    model code is available in the [appendix](#appendix-a) and the full
    project in the form of a reproducible pipeline, including the
    Rmarkdown source for this manuscript, is available at
    [https://github.com/jsks/ma-thesis](https://github.com/jsks/ma-thesis).

@@

colnames(df) <- translate(colnames(df))
q <- summary_stats(df)

@@

```{r results = "hide", fig.width = 6, fig.cap = "\\label{fig:map} (ref:map)"}
map_data <- st_read("data/raw/cshapes_0.6/cshapes.shp") %>%
    filter(GWEYEAR == max(GWEYEAR)) %>%
    select(country_name = CNTRY_NAME, gwid = GWCODE)

@@

rwanda_idx <- filter(final.df, country_name == "Rwanda", year %in% c(2000, 2010)) %$%
    reduced_idx

rwanda_p <- signif(mean(theta[, rwanda_idx[1]] < theta[, rwanda_idx[2]]) * 100, 2)
```

Finally, to expand upon the country selection from figure
\ref{fig:theta} and provide a temporal comparison, figure
\ref{fig:tie} plots the $\theta$ estimates for 2010 versus 2000 by
country. Note, the statistical significance of differences
between years for a single country cannot strictly be judged by
overlapping credible intervals [@Schenker2001]. For example, Chile,
despite having overlapping credible intervals, still has a posterior
probability of approximately $`r chile_p`$ percent that the estimate
for 2000 is lower than 2010. In the case of Rwanda with even more
visual overlap, there is a $`r rwanda_p`$ percent posterior
probability that the estimate for 2010 is higher than 2000.

@@

beta <- extract(reg, "beta")
columns <- c("exec_constraints", colnames(X),
             sprintf("exec * %s", colnames(X)[interaction_idx]))

colnames(beta) <- translate(columns)

@@

```{r results = "hide"}
argentina <- filter(merged.df, country_name == "Argentina", year == 1989) %$%
    fmt(rgdpepc)

slovakia <- filter(merged.df, country_name == "Slovakia", year == 2003) %$%
    fmt(rgdpepc)

cmb <- filter(merged.df, country_name == "Cambodia", year == 1987) %>%
    select(pop, area_sqkm) %>%
    mutate(area_sqkm = fmt(area_sqkm),
           pop = fmt(pop))

india <- filter(merged.df, country_name == "India", year == 1976) %>%
    select(pop, area_sqkm, meanelev) %>%
    mutate(pop = fmt(pop),
           area_sqkm = fmt(area_sqkm),
           meanelev = fmt(meanelev))

georgia <- filter(merged.df, country_name == "Georgia", year == 2012) %$%
    fmt(meanelev)

russia <- filter(merged.df, country_name == "Russia", year == 2012) %$%
    fmt(meanelev)

thai <- filter(merged.df, country_name == "Thailand", year == 2002) %$%
    { fmt(discrimpop * 100) }

bhutan <- filter(merged.df, country_name == "Bhutan", year == 1989) %$%
    { fmt(discrimpop * 100) }

@@

alpha_s <- quantile(alpha, probs = c(0.1, 0.9)) %>% fmt
alpha_interval <- sprintf("[%s,%s]", alpha_s[1], alpha_s[2])
```

@@

```{r results = "hide"}
gdp_rwa <- filter(merged.df, country_name == "Rwanda", year == 2007) %$%
    fmt(rgdpepc)

gdp_phl <- filter(merged.df, country_name == "Philippines", year == 2008) %$%
    fmt(rgdpepc)

gdp_po <- filter(merged.df, country_name == "Portugal", year == 2009) %$%
    fmt(rgdpepc)
```

@@

## Stan Code {#appendix-a}

@@
